---
title: "Calorie & Weight Tracker"
author: "Akeem Lawrence"
date: "2025-04-28"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
---

# Overview

This project is an interactive calorie and weight tracker built in R.

The tracker allows a user to:

- Log meals and calorie intake
- Log body weight
- Save tracking data between R sessions
- View daily calorie and weight summaries
- Review historical logs
- Visualize calorie intake over time
- Visualize body-weight trends
- Optionally export tracking data to Google Sheets

Personal calorie and weight data are stored locally and can be excluded from version control when the project is published to GitHub.

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)

# Create the local data directory if it does not exist
data_dir <- "data"

if (!dir.exists(data_dir)) {
  dir.create(
    data_dir,
    recursive = TRUE,
    showWarnings = FALSE
  )
}

# Define file locations
calorie_file <- file.path(
  data_dir,
  "calorie_log.csv"
)

weight_file <- file.path(
  data_dir,
  "weight_log.csv"
)
```

# Load Tracking Data

When the tracker starts, existing calorie and weight records are loaded from local CSV files.

If no files exist yet, empty tracking tables are created automatically.

```{r load-data}
if (file.exists(calorie_file)) {

  calorie_log <- read_csv(
    calorie_file,
    show_col_types = FALSE
  ) %>%
    mutate(Date = as.Date(Date))

} else {

  calorie_log <- tibble(
    Date = as.Date(character()),
    Food = character(),
    Calories = numeric()
  )
}

if (file.exists(weight_file)) {

  weight_log <- read_csv(
    weight_file,
    show_col_types = FALSE
  ) %>%
    mutate(Date = as.Date(Date))

} else {

  weight_log <- tibble(
    Date = as.Date(character()),
    Weight = numeric()
  )
}
```

# Save Tracking Data

This function saves calorie and weight entries to local CSV files.

```{r save-data}
save_logs <- function() {

  # Ensure the data directory exists before attempting to save
  if (!dir.exists(data_dir)) {
    dir.create(
      data_dir,
      recursive = TRUE,
      showWarnings = FALSE
    )
  }

  write_csv(
    calorie_log,
    calorie_file
  )

  write_csv(
    weight_log,
    weight_file
  )
}
```

# Daily Summary

The daily summary calculates total calories logged for the current day and displays the most recent weight entry, if one is available.

```{r daily-summary}
show_summary <- function() {

  today_date <- Sys.Date()

  todays_meals <- calorie_log %>%
    filter(Date == today_date)

  todays_total_calories <- sum(
    todays_meals$Calories,
    na.rm = TRUE
  )

  todays_weight <- weight_log %>%
    filter(Date == today_date)

  cat("\n")
  cat("=================================\n")
  cat("        DAILY SUMMARY\n")
  cat("=================================\n")

  cat(
    "Date:",
    format(today_date, "%B %d, %Y"),
    "\n"
  )

  cat(
    "Calories logged:",
    todays_total_calories,
    "kcal\n"
  )

  cat(
    "Meals logged:",
    nrow(todays_meals),
    "\n"
  )

  if (nrow(todays_weight) > 0) {

    latest_weight <- tail(
      todays_weight$Weight,
      1
    )

    cat(
      "Latest weight:",
      latest_weight,
      "lbs\n"
    )

  } else {

    cat(
      "Weight: No weight logged today\n"
    )
  }

  cat("=================================\n\n")
}
```

# Log a Meal

The meal logger records the current date, food or meal description, and calorie amount. The entry is immediately saved to the local calorie log.

```{r log-meal}
log_meal <- function() {

  food <- trimws(
    readline(
      "Enter food or meal: "
    )
  )

  if (food == "") {

    cat(
      "Food name cannot be empty.\n"
    )

    return()
  }

  calories <- suppressWarnings(
    as.numeric(
      readline(
        "Enter calories: "
      )
    )
  )

  if (
    is.na(calories) ||
    calories < 0
  ) {

    cat(
      "Please enter a valid calorie amount.\n"
    )

    return()
  }

  new_entry <- tibble(
    Date = Sys.Date(),
    Food = food,
    Calories = calories
  )

  calorie_log <<- bind_rows(
    calorie_log,
    new_entry
  )

  save_logs()

  cat(
    "\nMeal logged successfully.\n"
  )

  show_summary()
}
```

# Log Body Weight

The weight logger records the user's current body weight and saves the value to the weight history.

```{r log-weight}
log_weight <- function() {

  weight <- suppressWarnings(
    as.numeric(
      readline(
        "Enter weight in pounds: "
      )
    )
  )

  if (
    is.na(weight) ||
    weight <= 0
  ) {

    cat(
      "Please enter a valid weight.\n"
    )

    return()
  }

  new_entry <- tibble(
    Date = Sys.Date(),
    Weight = weight
  )

  weight_log <<- bind_rows(
    weight_log,
    new_entry
  )

  save_logs()

  cat(
    "\nWeight logged successfully.\n"
  )

  show_summary()
}
```

# View Historical Logs

These functions allow the user to review previously recorded calorie and weight entries.

```{r view-logs}
view_calorie_log <- function() {

  if (nrow(calorie_log) == 0) {

    cat(
      "No calorie data has been logged yet.\n"
    )

    return()
  }

  print(
    calorie_log %>%
      arrange(desc(Date))
  )
}


view_weight_log <- function() {

  if (nrow(weight_log) == 0) {

    cat(
      "No weight data has been logged yet.\n"
    )

    return()
  }

  print(
    weight_log %>%
      arrange(desc(Date))
  )
}
```

# Visualize Calorie Intake

Daily calorie totals are aggregated and visualized over time.

```{r calorie-plot}
plot_calories <- function() {

  if (nrow(calorie_log) == 0) {

    cat(
      "No calorie data available to plot.\n"
    )

    return()
  }

  daily_calories <- calorie_log %>%
    group_by(Date) %>%
    summarise(
      Daily_Calories = sum(
        Calories,
        na.rm = TRUE
      ),
      .groups = "drop"
    )

  calorie_plot <- ggplot(
    daily_calories,
    aes(
      x = Date,
      y = Daily_Calories
    )
  ) +
    geom_line() +
    geom_point(size = 2) +
    labs(
      title = "Daily Calorie Intake",
      x = "Date",
      y = "Calories"
    ) +
    theme_minimal()

  print(calorie_plot)
}
```

# Visualize Weight Trends

Body-weight entries are visualized chronologically to show changes over time.

```{r weight-plot}
plot_weight <- function() {

  if (nrow(weight_log) == 0) {

    cat(
      "No weight data available to plot.\n"
    )

    return()
  }

  weight_plot <- ggplot(
    weight_log,
    aes(
      x = Date,
      y = Weight
    )
  ) +
    geom_line() +
    geom_point(size = 2) +
    labs(
      title = "Body Weight Over Time",
      x = "Date",
      y = "Weight (lbs)"
    ) +
    theme_minimal()

  print(weight_plot)
}
```

# Optional Google Sheets Export

Tracking data can optionally be exported to Google Sheets.

Google authentication is only triggered when this function is intentionally run.

```{r google-sheets-export}
export_to_google_sheets <- function() {

  if (
    !requireNamespace(
      "googlesheets4",
      quietly = TRUE
    )
  ) {

    cat(
      paste0(
        "Install googlesheets4 first with:\n",
        'install.packages("googlesheets4")\n'
      )
    )

    return()
  }

  googlesheets4::gs4_auth()

  sheet_name <- paste0(
    "Calorie and Weight Tracker - ",
    Sys.Date()
  )

  sheet <- googlesheets4::gs4_create(
    sheet_name
  )

  googlesheets4::sheet_write(
    calorie_log,
    ss = sheet,
    sheet = "Calorie Log"
  )

  googlesheets4::sheet_write(
    weight_log,
    ss = sheet,
    sheet = "Weight Log"
  )

  cat(
    "\nData exported to Google Sheets successfully.\n"
  )
}
```

# Interactive Tracker

The main application provides a menu for entering data, viewing summaries, reviewing historical records, visualizing trends, and exporting data.

```{r tracker-app}
run_tracker <- function() {

  repeat {

    cat("\n")
    cat("=================================\n")
    cat("     CALORIE & WEIGHT TRACKER\n")
    cat("=================================\n")
    cat("1: Log Meal\n")
    cat("2: Log Weight\n")
    cat("3: View Daily Summary\n")
    cat("4: View Calorie Log\n")
    cat("5: View Weight Log\n")
    cat("6: Plot Calories\n")
    cat("7: Plot Weight\n")
    cat("8: Export to Google Sheets\n")
    cat("9: Exit\n")
    cat("=================================\n")

    choice <- readline(
      "Enter choice: "
    )

    if (choice == "1") {

      log_meal()

    } else if (choice == "2") {

      log_weight()

    } else if (choice == "3") {

      show_summary()

    } else if (choice == "4") {

      view_calorie_log()

    } else if (choice == "5") {

      view_weight_log()

    } else if (choice == "6") {

      plot_calories()

    } else if (choice == "7") {

      plot_weight()

    } else if (choice == "8") {

      export_to_google_sheets()

    } else if (choice == "9") {

      cat(
        "\nTracker closed. Goodbye!\n"
      )

      break

    } else {

      cat(
        "Invalid selection. Please choose 1 through 9.\n"
      )
    }
  }
}
```

# Running the Application

The interactive menu is intentionally not started automatically when this RMarkdown document is rendered.

After running the code chunks in RStudio, start the application from the Console with:

```r
run_tracker()
```

# Data Privacy

The tracker stores calorie and body-weight records locally in the `data` directory.

For a public GitHub repository, these personal CSV files should be excluded using `.gitignore`.

# Future Development

Potential future improvements include:

- Protein, carbohydrate, and fat tracking
- Daily calorie and macronutrient goals
- Weekly and monthly summaries
- Rolling weight averages
- Additional nutrition analytics
- Database-backed storage
- Interactive dashboard
- Web or mobile interface
- Cloud deployment